---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Author: Daniel Kinpara  
Date: April, 2015  

```{r configuration, include=FALSE}
Sys.setlocale("LC_TIME", "English")
Sys.setenv(tz = "UTC")
options(scipen = 999)
```
### 1. Loading and preprocessing the data

The first step was to read the file **activity.zip** and to save it in the object *dados*. The ZIP file must be placed in the same folder of this Markdown document in order to run the embedded code. The variable **date** was converted to a *date* class.

```{r echo=TRUE, warning=FALSE}
dados <- read.csv(bzfile("repdata-data-StormData.csv.bz2"), header = TRUE, na.strings = "NA",
                  colClasses = c("NULL", "character", "NULL", "character", "NULL", "NULL",
                                 "factor", "factor", "NULL", "NULL", "NULL", "character",
                                 rep("NULL", 10), "numeric", "numeric", "numeric",
                                 "character", "numeric", "character", rep("NULL", 8), "numeric"))
str(dados)

```
  
### 2. What is mean total number of steps taken per day?  

The code below created a factor variable named **dia** that was used to split the dataset.