---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
---
  
---
  
#Major Weather Events in the US
##and its Economic and Health Impacts, 1950-2011

Author: Daniel Kinpara  
Date: April, 2015  

```{r configuration, include=FALSE}
Sys.setlocale("LC_TIME", "English")
Sys.setenv(tz = "UTC")
options(digits = 2)
library(stringr)
```
### 1. Synopsis

The objective of this study is to determine the five most harmful types of weather events in the United States among the 48 listed in the Storm Data Event Table listed by the National Weather Service (NWS). The criterias used to rank the harmfulness are the injuries and fatalities caused to the population and the economic damages due to weather events. The study will comprehend the period of 1950 to 2011. It will use the National Oceanic & Atmospheric Administration's (NOAA) Storm Data database. For more details of how injuries and fatalities are accounted and economic damages calculated, see the document **Storm Data Preparation** available at [this website](http://www.nws.noaa.gov/directives/).

---

### 2. Data processing

#### 2.1 Loading dataset

The dataset is available in a compressed format (*bz2* extension) from [this link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2). The file size is 47MB, which consumes time and memory resources to be complete loaded. In order to speed up the loading process and to use less resources, only nine variables were read from the data file, as listed below:

1. **BGN-DATE** (character): date and time of the event occurrence ;
2. **EVTYPE** (character): 985 event types (when converted to class factor);
3. **FATALITIES** (numeric): number of deaths, direct or indirect related to the event;
4. **INJURIES** (numeric): number of injuried people;
5. **PROPDMG** (numeric): mantissa of the number of damages to properties;
6. **PROPDMGEXP** (character): exponent of the previous number, coded as:
    - "K", thousand US dollars;
    - "M", million US dollars;
    - "B", billion US dollars;
7. **CROPDMG** (numeric): mantissa of the crop damages;
8. **CROPDMGEXP** (character): exponent of the previous number, same coding of PROPDMGEXP;
9. **REFNUM** (numeric): unique index number.

The following code loads the dataset into **dados**, skipping unused columns. Missing data was assigned *NA*.

```{r loading, echo=TRUE, warning=FALSE, cache=TRUE}
dados <- read.csv(bzfile("repdata-data-StormData.csv.bz2"),
                  header = TRUE, na.strings = "NA",
                  colClasses = c("NULL", "character", rep("NULL", 5),
                                 "character", rep("NULL", 4), rep("NULL", 10),
                                 "numeric","numeric", "numeric", "character",
                                 "numeric", "character",rep("NULL", 8),
                                 "numeric"))
```
  
The dataset structure is:

```{r echo=TRUE}
str(dados)
```

#### 2.2 Pre-processing data

Some pre-processing was necessary in order to reduce the number of variables and to allow numerical calculation. Date was extracted from **BGN-DATE** and stored into the variable **tempo**, a class date variable. **PROPDMG** and **PROPDMGEXP** were combined into the numeric variable **urbano**. **CROPDMG** and **CROPDMGEXP** were combined into the numeric variable **rural**. Both were set to the same scale of dollars.
  
```{r smoothing, echo=TRUE, warning=FALSE, cache=TRUE}
dados <- transform(dados, tempo = as.Date(dados$BGN_DATE, "%m/%d/%Y"),
                   urbano = (dados$PROPDMG * 1e9 * (dados$PROPDMGEXP == "B") +
                             dados$PROPDMG * 1e6 * (dados$PROPDMGEXP == "M") +
                             dados$PROPDMG * 1e3 * (dados$PROPDMGEXP == "K") +
                             dados$PROPDMG * (dados$PROPDMGEXP == "")),
                   rural = (dados$CROPDMG * 1e9 * (dados$CROPDMGEXP == "B") +
                            dados$CROPDMG * 1e6 * (dados$CROPDMGEXP == "M") +
                            dados$CROPDMG * 1e3 * (dados$CROPDMGEXP == "K") +
                            dados$CROPDMG * (dados$CROPDMGEXP == "")))
dados <- dados[, c(2:4, 9:12)]
```
  
The new dataset structure is:

```{r echo=TRUE}
str(dados)
```
  
The next step was to clean **EVTYPE**. There are 985 different event types in the database. The non-standard events were re-classified into the 48 listed event types defined by the NWS. In order to do that, each observation was compared to the event types in the official list. If there's a match, the event was assigned as the respective match. Otherwise, the event was assigned as **NA**. The content of **EVTYPE** had the trailing spaces removed and the letters capitalized before matching.
  
```{r events, echo=TRUE, cache=TRUE}
eventTypes <- c("ASTRONOMICAL LOW TIDE", "AVALANCHE", "BLIZZARD", "COASTAL FLOOD",
                "COLD/WIND CHILL", "DEBRIS FLOW", "DENSE FOG", "DENSE SMOKE",
                "DROUGHT", "DUST DEVIL", "DUST STORM", "EXCESSIVE HEAT",
                "EXTREME COLD/WIND CHILL", "FLASH FLOOD", "FLOOD", "FROST/FREEZE",
                "FUNNEL CLOUD", "FREEZING FOG", "HAIL", "HEAT", "HEAVY RAIN",
                "HEAVY SNOW", "HIGH SURF", "HIGH WIND", "HURRICANE (TYPHOON)",
                "ICE STORM", "LAKE-EFFECT SNOW", "LAKESHORE FLOOD", "LIGHTNING",
                "MARINE HAIL", "MARINE HIGH WIND", "MARINE STRONG WIND", 
                "MARINE THUNDERSTORM WIND", "RIP CURRENT", "SEICHE", "SLEET",
                "STORM SURGE/TIDE", "STRONG WIND", "THUNDERSTORM WIND", "TORNADO",
                "TROPICAL DEPRESSION", "TROPICAL STORM", "TSUNAMI", "VOLCANIC ASH",
                "WATERSPOUT", "WILDFIRE", "WINTER STORM", "WINTER WEATHER")
dados <- transform(dados, evento = factor(rep("NA", 902297), levels = eventTypes))
for(i in 1:48){
        match <- grep(eventTypes[i], str_trim(toupper(dados$EVTYPE)))
        dados$evento[match] <- eventTypes[i]
}
```

```{r missing, echo=TRUE}
listNoMatch <- is.na(dados$evento)
noMatch <- sum(listNoMatch)
noMatchDMG <- round(with(dados, sum(rural[listNoMatch] + urbano[listNoMatch]) / 
                    sum(rural + urbano) * 100))
noMatchHTH <- round(with(dados, sum(FATALITIES[listNoMatch] + INJURIES[listNoMatch]) /
                           sum(FATALITIES + INJURIES) * 100))
```
  
After matching, **`r format(noMatch, big.mark = ",", scientific = FALSE)`** observations remained not classified and it was assigned **NA** to them. This represents **`r round(noMatch / 902297 * 100)`%** of the total observations, **`r noMatchDMG`%** of the total damages (properties + crops) and **`r noMatchHTH`%** of the total health impacts. They were not used for further computation of impacts. 
  
The final structure of the data is:  
  
```{r echo=TRUE}
dados <- dados[!listNoMatch,2:8]
str(dados)
```
  
---
  
### 3. Results

The data used in the analysis comprehend **`r format(length(dados$REFNUM), big.mark = ",", scientific = FALSE)`** observations of weather events that took place from **`r format(min(dados$tempo), "%m/%d/%Y")`** to **`r format(max(dados$tempo), "%m/%d/%Y")`**.
  
Along this period, weather events were responsible for **US$ `r format(sum(dados$urbano), big.mark = ",", scientific = FALSE)`** in properties damages and **US$ `r format(sum(dados$rural), big.mark = ",", scientific = FALSE)`** in crops loses. In terms of health, **`r format(sum(dados$INJURIES), big.mark = ",", scientific = FALSE)`** inhabitants were injured, directly or indirectly,  due to weather events. A total of **`r format(sum(dados$FATALITIES), big.mark = ",", scientific = FALSE)`** souls were lost.

#### 3.1 Economic impacts

The following code produced the graph of five events that caused more economic damages. Since the impacts have two categories, a panel with three graphs was presented. The graphs computed the properties damages, the crops damages, and the sum of both. It's important to notice that the graph *Total* was the sum of the overall economic damages due to any weather event. It didn't represent the simple sum of the graph *Properties* and graph *Crops*. The graphs presented the damage in terms of billion of dollars.
  
```{r economic, echo=TRUE, fig.height=4}
dados <- transform(dados, economico = (urbano + rural))

resUrbano <- with(dados, tapply(urbano, evento, sum))
ord <- order(sapply(resUrbano, "[", 1), decreasing = TRUE)
nomes1 <- names(resUrbano[ord[1:5]])

resRural <- with(dados, tapply(rural, evento, sum))
ord <- order(sapply(resRural, "[", 1), decreasing = TRUE)
nomes2 <- names(resRural[ord[1:5]])

resEcon <- with(dados, tapply(economico, evento, sum))
ord <- order(sapply(resEcon, "[", 1), decreasing = TRUE)
nomes3 <- names(resEcon[ord[1:5]])

x1 <- c(1:5)
y1 <- as.numeric(resUrbano[ord[1:5]])/1e9
y2 <- as.numeric(resRural[ord[1:5]])/1e9
y3 <- as.numeric(resEcon[ord[1:5]])/1e9

par(mfrow = c(1, 3))
barplot(y1, col = rainbow(5), ylim = c(0, 200), axes = TRUE,
        ylab = "billion US$", main = "Properties")
legend("topright", nomes1, fill = rainbow(5), cex = 0.8)
barplot(y2, col = rainbow(5), ylim = c(0, 200), axes = TRUE,
        ylab = "billion US$", main = "Crops")
legend("topright", nomes2, fill = rainbow(5), cex = 0.8)
barplot(y3, col = rainbow(5), ylim = c(0, 200), axes = TRUE,
        ylab = "billion US$", main = "Total")
legend("topright", nomes3, fill = rainbow(5), cex = 0.8)
```
  
Observing the *Total* graph, flood was the major cause of economic damages by far. More than twice the damage caused by tornados. The total economic damage caused by flood was **US$ `r format(y3[1], big.mark = ",", scientific = FALSE)`** billion dollars. Tornado accounted for **US$ `r format(y3[2], big.mark = ",", scientific = FALSE)`** billion dollars.
  
Damage to properties overaome damage to crops. Crop damages corresponded to only **`r with(dados, sum(rural)/sum(economico)) * 100`%** of total economic damages. Properties indeed were more affected by flood and tornado. However, crops suffered more from hail and drought. Crops seemed to be more damaged by events that have phisyological effects than events that have mechanical effects.
  
####3.2 Health impacts
  
The next code  produced the graph of five events that caused more health damages. Health damages were classified as injuries or fatalities. So, as the economic impact graphs, there were three graphs too. The graphs computed the injuries to people, the deaths, and the sum of both in terms of thousand of inhabitants. The graph *Total* was the sum of the overall injuries and deaths. It didn't represent the simple sum of the graph *Injuries* and graph *Fatalities*.
  
```{r health, echo=TRUE, fig.height=4}
dados <- transform(dados, populacao = (INJURIES + FATALITIES))

resInjuries <- with(dados, tapply(INJURIES, evento, sum))
ord <- order(sapply(resInjuries, "[", 1), decreasing = TRUE)
nomes1 <- names(resInjuries[ord[1:5]])

resFatalities <- with(dados, tapply(FATALITIES, evento, sum))
ord <- order(sapply(resFatalities, "[", 1), decreasing = TRUE)
nomes2 <- names(resFatalities[ord[1:5]])

resPop <- with(dados, tapply(populacao, evento, sum))
ord <- order(sapply(resPop, "[", 1), decreasing = TRUE)
nomes3 <- names(resPop[ord[1:5]])

x1 <- c(1:5)
y1 <- as.numeric(resInjuries[ord[1:5]])/1e3
y2 <- as.numeric(resFatalities[ord[1:5]])/1e3
y3 <- as.numeric(resPop[ord[1:5]])/1e3

par(mfrow = c(1, 3))
barplot(y1, col = rainbow(5), ylim = c(0, 100), ylab = "thousand inhabitants",
        main = "Injuries")
legend("topright", nomes1, fill = rainbow(5), cex = 0.8)
barplot(y2, col = rainbow(5), ylim = c(0, 100), ylab = "thousand inhabitants",
        main = "Fatalities")
legend("topright", nomes2, fill = rainbow(5), cex = 0.8)
barplot(y3, col = rainbow(5), ylim = c(0, 100), ylab = "thousand inhabitants",
        main = "Total")
legend("topright", nomes3, fill = rainbow(5), cex = 0.8)
```
  
As the graphs present, tornado was the major cause of injuries and fatalities. Tornado accounted for **`r format(y3[1], big.mark = ",", scientific = FALSE)`** thousand inahitants with health incidents. More than seven times the incidents caused by heat.  Heat accounted for **`r format(y3[2], big.mark = ",", scientific = FALSE)`** thousand inahitants with health incidents.

Different from the economic impacts, the four major causes of health impacts were the same for injuries and fatalities: tornado, heat, flood, and lightning. The only difference was that rip current caused more fatalities than thunderstorm wind. On the other hand, thunderstorm wind caused more injuries than rip current.

It's also noticeable that the proportion of injuries and fatalities was much higher in tornado events than heat events. While there was one fatality for each **`r format(round(y1[1]/y2[1]), big.mark = ",", scientific = FALSE)`** injuried individuals due to tornados, heat proportion was of one fatality for each **`r format(round(y1[2]/y2[2]), big.mark = ",", scientific = FALSE)`** injuried individuals. So, the event of heat was more likely to cause a higher percentage of deaths than the event of a tornado.

---
  
### 4. Final considerations

This analysis was quite superficial and it didn't explore all the possibilities presented by the data available. It's suggested further studies to encompass:
  
- the impact differences among the North American states;
- the change of the weather events severity along time;
- the mitigation of impacts due to a better understanding of the mechanism behind the effect of the weather event over crops and people;
- the relation between the event's duration and frequency and the impacts observerd;
- and the recurrence of severe weather events.

Of course, it's not an exhaustive list of possible studies, but they may present a solid base of longitudinal studies in order to stablish trends for the weather events. So, more strategies can be designed to reduce the material and life losses.
